---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  dpi = 1000,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%")
```

# urbnindicators

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL
v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![Codecov test
coverage](https://codecov.io/gh/UI-Research/urbnindicators/graph/badge.svg)](https://app.codecov.io/gh/UI-Research/urbnindicators)
<!-- badges: end -->

# Overview

**urbnindicators** aims to provide users with analysis-ready data from
the American Community Survey (ACS).

What you can access:

-   Hundreds of pre-computed variables, including percentages and
    the raw count variables used to produce them.

-   Or flexibly specify your own derived variables with a series of
    helper functions.

-   Margins of error for all variables--those direct from the API as
    well as derived variables--with correctly calculated pooled margins
    of error, per Census Bureau guidance.

-   Meaningful, consistent variable names--no more "B01003_001"; try 
    "total_population_universe" instead. (But if you're fond of the API's
    variable names, those are stored in the codebook as well for cross-referencing.)

-   A codebook that describes how each variable is calculated.

-   Data for multiple years and multiple states out of the box.

-   Supplemental measures, such as population density, that aren't
    available from the ACS.

-   Tools to aggregate or interpolate your data to different 
    geographies--along with correctly adjusted margins of error.

-   A local Shiny app to render your selected data and interactively explore,
    interpolate, and export those data to disk.


# Installation

Install the development version of `urbnindicators` from
[GitHub](https://github.com/) with:

```r
# install.packages("renv")
renv::install("UI-Research/urbnindicators")
```

You'll need a Census API key
([request one here](https://api.census.gov/data/key_signup.html)).
Set it once with:

```r
tidycensus::census_api_key("YOUR_KEY", install = TRUE)
```

Note that this package is under active development with frequent
updates--check to ensure you have the most recent version installed!

## Use with an AI agent

`urbnindicators` ships an [agent
skill](https://docs.claude.com/en/docs/agents-and-tools/agent-skills/overview)
that teaches coding agents how to use the package: how to discover variable
names rather than guess them, how to scope a query, how percentages and margins
of error are encoded, and how to aggregate to custom geographies. Register it
once, after installing the package:

```r
register_urbnindicators_skill()
```

That installs it for Claude Code. Pass `agent` to target others, or
`agent = "all"` for every supported agent:

```r
register_urbnindicators_skill(agent = c("claude", "codex", "gemini"))
```

| `agent` | Tool | How it loads |
|----|----|----|
| `"claude"` (default) | Claude Code, Claude Desktop | Native skill directory, loaded automatically when a request involves ACS data |
| `"codex"` | OpenAI Codex | Pointer added to `AGENTS.md` |
| `"gemini"` | Google Gemini CLI | Pointer added to `GEMINI.md` |
| `"agents"` | Cursor and other `AGENTS.md` adopters | Pointer added to `AGENTS.md` |

Only Claude discovers skill directories natively. For the others the skill files
are installed alongside the agent's configuration and a short pointer block is
added to its instructions file, so the full skill is read only when a task
actually calls for it rather than loaded into every conversation.

Use `scope = "project"` to install into the current repository instead of your
home directory, so collaborators inherit the skill. Re-run after upgrading the
package to pick up changes; re-running replaces the managed block and leaves the
rest of your instructions file untouched.

# Use

```{r, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(urbnthemes)
library(distributional)
library(ggdist)
set_urbn_defaults(style = "print")

library(urbnindicators)
```

## Discover Available Data

```{r, warning = FALSE, message = FALSE}
list_tables() |> head(10)
```

## Obtain Data

A single call to `compile_acs_data()` returns analysis-ready data
with pre-computed percentages, meaningful variable names, and
margins of error:

```{r, warning = FALSE, message = FALSE}
df = compile_acs_data(
  tables = "race",
  years = c(2019, 2024),
  geography = "county",
  states = "NJ")

df %>%
  select(1:10) %>%
  glimpse()
```

## Visualize Data

`compile_acs_data()` makes it easy to pull multiple years and
produce publication-ready visualizations:

```{r, warning = FALSE, message = FALSE}
plot_data = df %>%
  transmute(
    county_name = NAME %>% str_remove(" County, New Jersey"),
    race_personofcolor_percent,
    race_personofcolor_percent_M,
    data_source_year = factor(data_source_year))

state_averages = plot_data %>%
  summarize(
    .by = data_source_year,
    mean_pct = mean(race_personofcolor_percent)) %>%
  arrange(data_source_year) %>%
  pull(mean_pct)

## order counties by 2019 value for the dumbbell plot
county_order = plot_data %>%
  filter(data_source_year == "2019") %>%
  arrange(race_personofcolor_percent) %>%
  pull(county_name)

plot_data = plot_data %>%
  mutate(county_name = factor(county_name, levels = county_order))

dumbbell_data = plot_data %>%
  pivot_wider(
    id_cols = county_name,
    names_from = data_source_year,
    values_from = race_personofcolor_percent,
    names_prefix = "year_")

ggplot() +
  geom_segment(
    data = dumbbell_data,
    aes(
      x = county_name,
      y = year_2019,
      yend = year_2024),
    color = palette_urbn_main[7],
    linewidth = 1) +
  ggdist::stat_gradientinterval(
    data = plot_data,
    aes(
      x = county_name,
      ydist = distributional::dist_normal(
        race_personofcolor_percent,
        race_personofcolor_percent_M / 1.645),
      color = data_source_year),
    point_size = 2,
    .width = .95) +
  geom_hline(
    yintercept = state_averages[1],
    linetype = "dashed",
    color = palette_urbn_main[1]) +
  geom_hline(
    yintercept = state_averages[2],
    linetype = "dashed",
    color = palette_urbn_main[2]) +
  annotate(
    "text",
    y = state_averages[1] - .15,
    x = 21.5,
    label = "State mean (2019)",
    fontface = "bold.italic",
    color = palette_urbn_main[1],
    size = 9 / .pt,
    hjust = 0,
    nudge_y = .01) +
  annotate(
    "text",
    y = state_averages[2] + .01,
    x = 21.5,
    label = "State mean (2024)",
    fontface = "bold.italic",
    color = palette_urbn_main[2],
    size = 9 / .pt,
    hjust = 0,
    nudge_y = .01) +
  labs(
    title = "All NJ Counties Experienced Racial Diversification from 2019 to 2024",
    subtitle = paste0("Share of population who are people of color, by county, 2019-2024
Confidence intervals are presented around each point but are extremely small"),
    x = "",
    y = "Share of population who are people of color") +
  scale_x_discrete(expand = expansion(mult = c(.03, .04))) +
  scale_y_continuous(
    breaks = c(0, .25, .50, .75, 1.0),
    limits = c(0, .75),
    labels = scales::percent) +
  coord_flip() +
  theme_urbn_print()
```

## Supported Geographies

`compile_acs_data()` returns five-year ACS estimates for any geography
accepted by `tidycensus::get_acs()`. Estimates are available for **2009
and later** (the first year of five-year ACS data); `geography` defaults
to `"county"` and `years` to the most recent release.

| `geography` | `states` argument | Notes |
|---|---|---|
| `"us"`, `"region"`, `"division"`, `"cbsa"`, `"zcta"`, `"urban area"` | Ignored | Returned nationally |
| `"state"` | Optional (filters) | States and DC |
| `"county"` (default), `"place"`, `"county subdivision"` | Recommended | Omitting `states` pulls every state, which is slow |
| `"tract"` | Required | Census tracts |
| `"block group"` | Required | 2013 and later only. Only a subset of tables is published at this geography (see `list_tables(geography = "block group")`). Estimates carry large margins of error |

Census blocks are not supported, as the ACS
publishes no block-level data. For `"county"`, `"tract"`, and
`"block group"`, you can also pass `counties` (five-digit FIPS codes)
to limit the pull to specific counties. See `vignette("acs-background")`
for more on geographies, including boundary changes across the 2020
census.

## Custom Geographies

ACS data are available for standard geographies (tracts, counties,
states, etc.), but many analyses require non-standard areas like
neighborhoods, school zones, or planning districts.
`interpolate_acs()` aggregates source data to
any user-defined geography, properly re-deriving percentages and
propagating margins of error:

```{r, warning = FALSE, message = FALSE}
dc_tracts = compile_acs_data(
  tables = "snap",
  years = 2024,
  geography = "tract",
  states = "DC",
  spatial = TRUE)

## assign each tract to a quadrant based on its centroid
dc_tracts = dc_tracts %>%
  mutate(
    centroid = sf::st_centroid(geometry),
    lon = sf::st_coordinates(centroid)[, 1],
    lat = sf::st_coordinates(centroid)[, 2],
    quadrant = case_when(
      lon <  median(lon) & lat >= median(lat) ~ "NW",
      lon >= median(lon) & lat >= median(lat) ~ "NE",
      lon <  median(lon) & lat <  median(lat) ~ "SW",
      lon >= median(lon) & lat <  median(lat) ~ "SE")) %>%
  select(-centroid, -lon, -lat)

## aggregate tracts to quadrants
dc_quadrants = interpolate_acs(
  .data = dc_tracts,
  target_geoid = "quadrant")

dc_quadrants %>%
  sf::st_drop_geometry() %>%
  select(GEOID, snap_received_percent, snap_received_percent_M)
```

See `vignette("custom-geographies")` for more.

## Custom Derived Variables

Beyond the package's built-in tables, you can define your own
derived variables using the `define_*()` helpers and pass them
directly to `compile_acs_data()`. Your custom variables
automatically get codebook entries and correctly calculated margins of error:

```{r, warning = FALSE, message = FALSE}
df = compile_acs_data(
  tables = list(
    "snap",
    define_percent(
      numerator = "snap_universe",
      denominator = "snap_universe",
      subtract_from_numerator = "snap_received",
      output = "snap_not_received_percent")),
  years = 2024,
  geography = "county",
  states = "DC")

df %>%
  select(matches("snap.*percent")) %>%
  glimpse()
```

See `vignette("custom-derived-variables")` for detailed examples of
each of the `define_*()` helpers.

# Learn More

Check out the vignettes for additional details:

-   A package overview to help users [**Get
    Started**](articles/urbnindicators.html).

-   An interactive table of the package's
    [**Codebook**](articles/codebook.html).

-   An illustration of how [**Quantifying Survey
    Error**](articles/quantified-survey-error.html) can improve
    inference making.

-   A guide to defining [**Custom Derived
    Variables**](articles/custom-derived-variables.html) using the
    `define_*()` helpers.

-   You can re-create your indicators and their measures of error
    for [**Custom Geographies**](articles/custom-geographies.html).

-   See how you can spin up a [**Local Shiny app**](articles/view_acs_data.html)
    to explore, interpolate, and export your data.

# Credits

This package is built on top of `library(tidycensus)`, 
which provides the core functionality for
accessing the Census Bureau API. Learn more here: 
<https://walker-data.com/tidycensus/index.html>.

A number of staff in the Urban Institute's Housing and Communities
Division and Data Science and Technology team provided feedback, code reviewed,
and otherwise supported the development of this package.

## Use of Generative AI

This package was developed in part using agentic AI, under the
supervision of the author.
