---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# urbnindicators

<!-- badges: start -->
<!-- badges: end -->

The goal of `urbnindicators` is to provide users with a simple interface to generate a ready-for-analysis 
dataset of common social sciences measures. It is built on top of the `tidycensus` package and abstracts 
common query and cleaning tasks--such as identifying required variable codes to pass to `tidycensus::get_acs()`,
calculating percentages and other measures of interest that are derived from count estimates, and renaming variables
such that they are semantically-meaningful (but still programmatically functional). In addition to simplicity, it
includes a series of data quality checks to help ensure that measures are accurately computed and to provide users
with quantifications of measures' reliability (using coefficients of variation) rather than merely returning margins 
of error (which are often immediately dropped from analysis workflows without further consideration.

(aspirational)
`urbnindicators` also provides a set of related tools to facilitate common tasks when working with Census data 
(and ACS data in particular), such as calculating calculating measures of segregation, interpolating data across 
geographies and over time, and providing change-over-time measures. 


## Installation

Install the development version of urbnindicators from [GitHub](https://github.com/) with:

``` r
# install.packages("devtools")
devtools::install_github("UI-Research/urbnindicators")
```

This package is under active development with frequent updates--install an up-to-date version frequently!

## urbnindicators in action

```{r example}
library(urbnindicators)

acs_df = compile_acs_data(
  variables = NULL,
  years = c(2021, 2022),
  geography = "county",
  states = "NJ",
  counties = NULL,
  retain_moes = TRUE)

acs_df %>% 
  dplyr::select(c(
    NAME,
    data_source_year, 
    race_personofcolor_percent)) %>%
  dplyr::mutate(county_name = NAME %>% stringr::str_remove(" County, New Jersey")) %>%
  ggplot2::ggplot(
    ggplot2::aes(
      x = reorder(county_name, race_personofcolor_percent), 
      y = race_personofcolor_percent, 
      color = factor(data_source_year))) +
  ggplot2::geom_point() +
  ggplot2::labs(
    title = "Change in NJ Counties' Populations of Color",
    x = "County", 
    y = "Share of population who are people of color",
    color = "Year") +
  ggplot2::scale_y_continuous(
    breaks = c(0, .25, .50, .75, 1.0),
    labels = scales::percent) +
  ggplot2::coord_flip()
```
