---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  dpi = 1000,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%")
```

# urbnindicators

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL
v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)
[![Codecov test
coverage](https://codecov.io/gh/UI-Research/urbnindicators/graph/badge.svg)](https://app.codecov.io/gh/UI-Research/urbnindicators)

<!-- badges: end -->

# Overview

**urbnindicators** aims to provide users with analysis-ready data from
the American Community Survey (ACS).

With a single function call, you get:

-   Access to hundreds of standardized variables, such as percentages and
    the raw count variables used to produce them.

-   Margins of error and coefficients of variation for all
    variables--those direct from the API as well as derived variables.

-   Meaningful, consistent variable names.

-   A codebook that describes how each variable is calculated.

-   The built-in capacity to pull data for multiple years and multiple
    states.

-   Supplemental measures, such as population density, that aren't
    available from the ACS.

-   Built-in quality checks to help ensure that calculated variables 
    and measures of error are accurate. Plus some good, old-fashioned manual QC.
    That said--use at your own risk. We cannot and do not guarantee there aren't bugs.



# Installation

Install the development version of `urbnindicators` from
[GitHub](https://github.com/) with:

```r
# install.packages("renv")
renv::install("UI-Research/urbnindicators")
```

You'll want a Census API key
([request one here](https://api.census.gov/data/key_signup.html)).
Set it once with:

```r
tidycensus::census_api_key("YOUR_KEY", install = TRUE)
```

Note that this package is under active development with frequent
updates--check to ensure you have the most recent version installed!

# Use

```{r, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
library(dplyr)
library(tidyr)
library(stringr)
library(ggplot2)
library(urbnthemes)
library(distributional)
library(ggdist)
set_urbn_defaults(style = "print")

library(urbnindicators)
```

## Discover Available Data

```{r, warning = FALSE, message = FALSE}
list_tables() |> head(10)
```

```{r, warning = FALSE, message = FALSE}
list_variables() |> head(10)
```

## Obtain Data

A single call to `compile_acs_data()` returns analysis-ready data
with pre-computed percentages, meaningful variable names, and
margins of error:

```{r, warning = FALSE, message = FALSE}
df = compile_acs_data(
  tables = "race",
  years = 2024,
  geography = "county",
  states = "NJ")

glimpse(df) |> head(10)
```

## Visualize Data

`compile_acs_data()` makes it easy to pull multiple years and
produce publication-ready visualizations:

```{r, warning = FALSE, message = FALSE}
df = compile_acs_data(
  tables = "race",
  ## selecting 5-year ACS data from years with no measurement overlap
  years = c(2019, 2024),
  geography = "county",
  states = "NJ")

plot_data = df %>%
  transmute(
    county_name = NAME %>% str_remove(" County, New Jersey"),
    race_personofcolor_percent,
    race_personofcolor_percent_SE,
    data_source_year = factor(data_source_year))

state_averages = plot_data %>%
  summarize(
    .by = data_source_year,
    mean_pct = mean(race_personofcolor_percent)) %>%
  arrange(data_source_year) %>%
  pull(mean_pct)

## order counties by 2019 value for the dumbbell plot
county_order = plot_data %>%
  filter(data_source_year == "2019") %>%
  arrange(race_personofcolor_percent) %>%
  pull(county_name)

plot_data = plot_data %>%
  mutate(county_name = factor(county_name, levels = county_order))

dumbbell_data = plot_data %>%
  pivot_wider(
    id_cols = county_name,
    names_from = data_source_year,
    values_from = race_personofcolor_percent,
    names_prefix = "year_")

ggplot() +
  geom_segment(
    data = dumbbell_data,
    aes(
      x = county_name,
      y = year_2019,
      yend = year_2024),
    color = palette_urbn_main[7],
    linewidth = 1) +
  ggdist::stat_gradientinterval(
    data = plot_data,
    aes(
      x = county_name,
      ydist = distributional::dist_normal(
        race_personofcolor_percent,
        race_personofcolor_percent_SE),
      color = data_source_year),
    point_size = 2,
    .width = .95) +
  geom_hline(
    yintercept = state_averages[1],
    linetype = "dashed",
    color = palette_urbn_main[1]) +
  geom_hline(
    yintercept = state_averages[2],
    linetype = "dashed",
    color = palette_urbn_main[2]) +
  annotate(
    "text",
    y = state_averages[1] - .15,
    x = 21.5,
    label = "State mean (2019)",
    fontface = "bold.italic",
    color = palette_urbn_main[1],
    size = 9 / .pt,
    hjust = 0,
    nudge_y = .01) +
  annotate(
    "text",
    y = state_averages[2] + .01,
    x = 21.5,
    label = "State mean (2024)",
    fontface = "bold.italic",
    color = palette_urbn_main[2],
    size = 9 / .pt,
    hjust = 0,
    nudge_y = .01) +
  labs(
    title = "All NJ Counties Experienced Racial Diversification from 2019 to 2024",
    subtitle = paste0("Share of population who are people of color, by county, 2019-2024
Confidence intervals are presented around each point but are extremely small"),
    x = "",
    y = "Share of population who are people of color") +
  scale_x_discrete(expand = expansion(mult = c(.03, .04))) +
  scale_y_continuous(
    breaks = c(0, .25, .50, .75, 1.0),
    limits = c(0, .75),
    labels = scales::percent) +
  coord_flip() +
  theme_urbn_print()
```

# Learn More

Check out the vignettes for additional details:

-   A package overview to help users [**Get
    Started**](articles/urbnindicators.html).

-   An interactive version of the package's
    [**Codebook**](articles/codebook.html) so that prospective users can
    know what to expect.

-   A brief description of the package's [**Design
    Philosophy**](articles/design-philosophy.html) to clarify the
    use-cases that `urbnindicators` is built to support.

-   An illustration of how [**Quantifying Survey
    Error**](articles/quantified-survey-error.html) can improve
    inference making.

-   You can re-create your indicators and their measures of error
    for [**Custom Geographies**](articles/custom-geographies.html).
    Neighborhoods? Unincorporated counties? Start here.

# Credits

This package is built on top of and enormously indebted to
`library(tidycensus)`, which provides the core functionality for
accessing the Census Bureau API. For users who want additional
variables, `library(tidycensus)` exposes the entire range of
pre-tabulated variables available from the ACS and provides access to
ACS microdata and other Census Bureau datasets.

Learn more here: <https://walker-data.com/tidycensus/index.html>.
