---
output: github_document
editor_options: 
  markdown: 
    wrap: 72
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  collapse = TRUE,
  dpi = 1000,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%")
```

# urbnindicators

<!-- badges: start -->

[![Lifecycle:
experimental](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://lifecycle.r-lib.org/articles/stages.html#experimental)
[![License: GPL
v3](https://img.shields.io/badge/License-GPLv3-blue.svg)](https://www.gnu.org/licenses/gpl-3.0)

<!-- badges: end -->

The goal of `urbnindicators` is to provide users with analysis-ready
data from the American Community Survey (ACS). Here's what it offers as
compared to other solutions for working with ACS data:

-   Meaningful, consistent variable names.

-   Standardized variables, such as percentages, in addition to the raw
    count variables used to produce them.

-   A codebook that describes how each variable is calculated.

-   The built-in capacity to pull data for multiple years and multiple
    states.

-   Supplemental measures, such as population density, that aren't
    available from the ACS.

-   Built-in quality checks to help ensure that calculated variables are
    accurate.

And there's more in the works. Future updates will include:

-   Even more indicators.

-   Coefficients of variation for all variables to help measure estimate
    reliability.

-   Checks for variable alignment over time.

-   Support for inter-decadal geographic alignment.

## Installation

Install the development version of urbnindicators from
[GitHub](https://github.com/) with:

``` r
# install.packages("renv")
renv::install("UI-Research/urbnindicators")
```

This package is under active development with frequent updates--check to
ensure you have the most recent version installed!

## urbnindicators in action
```{r, include = FALSE, echo = FALSE, message = FALSE, warning = FALSE, eval = TRUE}
library(dplyr)
library(tidyr)
library(stringr)
library(urbnindicators)
library(ggplot2)
library(urbnthemes) 
set_urbn_defaults(style = "print")
```

<<<<<<< HEAD
```{r, include = F, echo = F, warning = F, mesage = F, eval = T}
library(dplyr)
library(stringr)
library(urbnindicators)
library(ggplot2)
```


```{r example}
# library(dplyr)
# library(stringr)
# library(urbnindicators)
# library(ggplot2)
=======
```{r example, fig.height = 5}
# library(dplyr)
# library(tidyr)
# library(stringr)
# library(urbnindicators)
# library(ggplot2)
# library(urbnthemes) 
# set_urbn_defaults(style = "print")
>>>>>>> covs-2

acs_df = compile_acs_data(
  variables = NULL,
  years = c(2017, 2022),
  geography = "county",
  states = "NJ",
  counties = NULL,
  spatial = FALSE) %>% 
  transmute(
    county_name = NAME %>% str_remove(" County, New Jersey"), 
    race_personofcolor_percent, 
    data_source_year)

state_averages = acs_df %>%
  group_by(data_source_year) %>%
  summarize(mean_race_personofcolor_percent = mean(race_personofcolor_percent)) %>%
  arrange(data_source_year) %>%
  pull()

dumbbell_data = acs_df %>% 
  pivot_wider(
    names_from = data_source_year, 
    values_from = race_personofcolor_percent, 
    names_prefix = "year_")

acs_df %>% 
  ggplot() +
    geom_segment(
      data = dumbbell_data,
      aes(
        x = reorder(county_name, year_2017),
        y = year_2017,
        yend = year_2022),
      color = palette_urbn_main[7],
      linewidth = 1) +
    geom_point(
      aes(
      x = reorder(county_name, race_personofcolor_percent), 
      y = race_personofcolor_percent, 
      color = factor(data_source_year))) +
    annotate(
      "text", 
      y = .31,  
      x = 21.5, 
      label = "State mean (2017)", 
      fontface = "bold.italic", 
      color = palette_urbn_main[1]) +
    annotate(
      "text", 
      y = .47,  
      x = 21.5, 
      label = "State mean (2022)", 
      fontface = "bold.italic", 
      color = palette_urbn_main[2]) +
    geom_hline(yintercept = state_averages[1], linetype = "dashed", color = palette_urbn_main[1]) +
    geom_hline(yintercept = state_averages[2], linetype = "dashed", color = palette_urbn_main[2]) +
    labs(
      title = "Change in NJ Counties' Racial Composition, 2017 to 2022",
      subtitle = paste0(
        "On average, counties in NJ saw a ", 
        round((state_averages[2] - state_averages[1]), digits = 3) * 100, 
        " pecentage point increase in the share of the population who are people of color from 2017 to 2022.") %>%
        str_wrap(120), 
      x = "County", 
      y = "Share of population who are people of color",
      color = "Year") +
    scale_x_discrete(expand = expansion(mult = c(.03, 0.04))) +
    scale_y_continuous(
      breaks = c(0, .25, .50, .75, 1.0),
      limits = c(0, .75),
      labels = scales::percent) +
    coord_flip()
```
