---
title: "ACS Table Structures"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{ACS Table Structures}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>")
```

```{r setup}
library(urbnindicators)
library(dplyr)
library(stringr)
library(tidycensus)
library(reactable)
```

This vignette explains how `urbnindicators` translates raw ACS
data into analysis-ready output. It covers:

1.  How ACS tables are structured at the Census Bureau
2.  How the package selects and renames variables
3.  How percentages are calculated
4.  What the output looks like and how to interpret it

## How ACS Tables Are Structured

The Census Bureau organizes ACS data into thousands of **tables**,
each identified by a code like `B22003` or `B01001`. Each table
contains a set of numbered variables---`B22003_001`, `B22003_002`,
etc.---that report counts for specific subgroups.

Tables have a hierarchical label structure. For example, table
B22003 (Receipt of Food Stamps/SNAP) looks like:

| Variable | Label | Role |
|---|---|---|
| B22003_001 | Total | Universe (all households) |
| B22003_002 | Received Food Stamps/SNAP | Subgroup count |
| B22003_003 | ... with 1 or more persons with a disability | Sub-subgroup |
| B22003_004 | ... with no persons with a disability | Sub-subgroup |
| B22003_005 | Did not receive Food Stamps/SNAP | Subgroup count |

The first variable (`_001`) is always the table **total**, or
universe---the count of all observations to which the table
applies. Subsequent variables break down the total into
subcategories, sometimes nested several levels deep.

### Table prefixes

The Census Bureau uses prefixes to distinguish table types:

-   **B** (Detailed Tables): The most granular tabulations,
    with the full set of categories.
-   **C** (Collapsed Tables): Simplified versions of B tables
    with fewer categories. A given C table typically mirrors a
    B table but groups some subcategories together.
-   **S** (Subject Tables): Pre-computed summaries, often
    including percentages and medians. These are not used by
    `urbnindicators` directly; the package computes its own
    percentages from B and C tables.

`urbnindicators` works with B and C tables. You can browse the
full Census Bureau variable catalog with `get_acs_codebook()`:

```{r}
get_acs_codebook() %>%
  filter(table == "B22003") %>%
  head(6)
```

## Two Paths: Registered Tables and Auto-Processed Tables

`compile_acs_data()` supports two kinds of table inputs, each
processed differently.

### Registered tables

These are tables that the package has pre-built definitions for.
They have curated variable names, carefully chosen denominators,
and tested percentage calculations. Use `list_tables()` to see
what's available:

```{r}
list_tables()
```

When you request a registered table by name (e.g.,
`tables = "snap"`), the package knows exactly which ACS variables
to fetch and how to calculate derived measures from them.

### Auto-processed tables

Any valid ACS table code can be passed directly (e.g.,
`tables = "B25070"`). The table is **auto-processed**: all of its
variables are fetched, names are generated algorithmically from
the Census Bureau's labels, and percentages are computed based on
the table's label hierarchy. This holds even when a registered
table covers the same code — `tables = "B22003"` returns the
auto-processed table, while `tables = "snap"` returns the
registered version. (Requesting both in the same call returns the
registered version and drops the auto-processed one with a
warning, since the two would otherwise fetch duplicate
variables.)

Auto-processed tables are a flexible route for the
thousands of ACS tables that `urbnindicators` doesn't have
pre-built definitions for. The tradeoff is that variable names
tend to be longer and less polished (they're derived mechanically
from Census labels).

## How Variables Are Selected and Renamed

### Registered tables

For registered tables, the package maintainers have manually identified
constructs that are widely used in analyses, vetted how to calculate these constructs,
and simplified the resulting variable names to follow a concise and consistent structure:

| ACS code | `urbnindicators` name |
|---|---|
| B22003_001 | `snap_universe` |
| B22003_002 | `snap_received` |

These names follow the pattern:
`[concept]_[subconcept]_[characteristic]_[metric]`.

Some registered tables use **pattern-based selection** instead of
listing variables one by one. For example, the `sex_by_age` table
selects all variables matching the pattern `B01001_`, which pulls
every variable in table B01001 and renames them by parsing the
Census Bureau's hierarchical labels into snake_case.

### Auto-processed tables

For auto-processed tables, variable names are generated by
converting the Census Bureau's `concept` and `label` fields to
snake_case. For example, a variable from table B25070 with
concept "GROSS RENT AS A PERCENTAGE OF HOUSEHOLD INCOME IN THE
PAST 12 MONTHS" and label "Less than 10.0 percent" becomes:

```
gross_rent_as_a_percentage_household_income_in_past_12_months_less_than_10_0_pct
```

(Common connector words like "of" and "the" are collapsed, and a
trailing "percent" becomes `pct` so that computed percentage
variables--which end in `_percent`--remain distinguishable from raw
ACS variables that happen to measure percentages.)

These names are functional but verbose. To see what names a
table will produce before pulling data, you can use
`get_acs_codebook()`:

```{r}
get_acs_codebook(table = "B25070")
```

## How Percentages Are Calculated

### Registered tables

Registered tables define their percentages explicitly. Each
table's definitions specify which variable(s) form the numerator,
which form the denominator, and what the output variable should
be called. For example, `snap_received_percent` is documented in
the codebook as:

```
Numerator = snap_received (B22003_002). Denominator = snap_universe (B22003_001).
```

Some percentage calculations are more complex. For example,
`disability_percent` sums numerous sex-by-age-by-disability
variables into a single numerator, then divides by the table
universe.

All percentage variables end with `_percent`.

### Auto-processed tables

For auto-processed tables, the package builds a label
hierarchy tree from the Census Bureau's ACS codebook. Each
variable's "parent" is determined by its position in the
hierarchy:

| Variable | Label | Depth | Parent |
|---|---|---|---|
| B25070_001 | Total | 1 | *(none---this is the universe)* |
| B25070_002 | Less than 10.0 percent | 2 | B25070_001 |
| B25070_003 | 10.0 to 14.9 percent | 2 | B25070_001 |

The `denominator` parameter on `compile_acs_data()` controls
which variable is used as the denominator for auto-computed
percentages:

-   **`"parent"`** (default): Each variable is divided by its
    nearest parent subtotal in the label hierarchy. For flat
    tables (like B25070, where all variables are direct children
    of the total), this is equivalent to dividing by the table
    total. For nested tables with subtotals, this produces
    within-group percentages.
-   **`"total"`**: Every variable is divided by the table total
    (`_001`), producing shares of the overall population
    regardless of hierarchy.
-   **A specific ACS variable code** (e.g., `"B25070_001"`):
    Every variable is divided by that variable.

To illustrate: consider a table with structure
Total > Male > Male 18-24, Male 25-34, ... and
Total > Female > Female 18-24, Female 25-34, ...

With `denominator = "parent"`, "Male 18-24" would be divided by
"Male" (its parent subtotal), yielding the share *of males* in
that age group. With `denominator = "total"`, "Male 18-24"
would be divided by "Total", yielding the share *of all
adults*.

The package skips percentage computation for tables that contain
medians, aggregates, averages, or other non-count measures.

## Understanding the Output

`compile_acs_data()` returns a wide-format dataframe with one
row per geography per year. Here's what the columns look like:

```{r}
df = compile_acs_data(
  tables = "snap",
  years = 2024,
  geography = "county",
  states = "DC")

names(df)
```

### Column types

The output contains several types of columns, identifiable by
their naming conventions:

| Pattern | Type | Example | Description |
|---|---|---|---|
| `GEOID` | Identifier | `"11001"` | FIPS code for the geography |
| `NAME` | Identifier | `"District of Columbia"` | Human-readable name |
| `data_source_year` | Identifier | `2024` | The ACS release year (end year of the 5-year window) |
| `*_universe` | Count | `snap_universe` | Table total / denominator population |
| *(no suffix)* | Count | `snap_received` | Raw ACS estimate |
| `*_percent` | Derived | `snap_received_percent` | Percentage (0--1 scale) |
| `*_M` | MOE | `snap_received_M` | Margin of error (90% confidence) |

Every raw count variable has a corresponding `_M` column with
its margin of error, sourced directly from the Census Bureau.
Every derived percentage variable also has a `_M` column, with
the margin of error calculated by `urbnindicators` using Census
Bureau-recommended formulas for pooling errors.

### The codebook

Every dataframe returned by `compile_acs_data()` carries a
codebook as an attribute. The codebook documents how each variable
was calculated:

```{r}
attr(df, "codebook")
```

The **definition** column is especially important: raw ACS
variables read `"This is a raw ACS estimate."`, while derived
variables spell out their inputs alongside the original ACS codes
(e.g., `"Numerator = snap_received (B22003_002). Denominator =
snap_universe (B22003_001)."`). The codebook's component
list-columns (`numerator_vars`, `denominator_vars`, and their
`_subtract_` counterparts) drive the pooled margin-of-error
calculations, so they always reflect the actual calculation.

### GEOIDs

The `GEOID` column contains FIPS codes whose structure depends
on the geography level:

| Geography | GEOID format | Example | Components |
|---|---|---|---|
| State | 2 digits | `"11"` | State |
| County | 5 digits | `"11001"` | State + County |
| Tract | 11 digits | `"11001001002"` | State + County + Tract |
| Place | 7 digits | `"1150000"` | State + Place |

You can parse these to join with other datasets. For example,
`substr(GEOID, 1, 2)` extracts the state code from any
sub-state geography.

## Discovering Available Data

`urbnindicators` provides several functions for exploring what
data is available:

| Function | Purpose | Requires API key? |
|---|---|---|
| `list_tables()` | Names of all registered tables | No* |
| `list_variables()` | Every variable mapped to its table | Yes |
| `get_acs_codebook()` | Browse all ACS variables with clean names | Yes |

\*`list_tables(geography = "block group")` also requires a key,
because block-group availability is read from the ACS variable
dictionary. The Census Bureau's variables metadata endpoint requires
an API key, so any function that loads the dictionary needs
`CENSUS_API_KEY` set (see
`tidycensus::census_api_key("YOUR_KEY", install = TRUE)`).

A typical discovery workflow:

1.  **Start with `list_tables()`** to see the registered tables.
2.  **Use `list_variables()` to preview** what a specific table
    will produce: `list_variables() %>% filter(table == "snap")`.
3.  **If your topic isn't covered**, search the full ACS catalog:
    `get_acs_codebook() %>% filter(str_detect(variable_clean, "poverty"))`.
4.  **Pass the table code directly** to `compile_acs_data()`:
    `compile_acs_data(tables = "B17001", ...)`.

## Registered Table Reference

The table below lists all pre-built tables available via
`list_tables()`.

```{r, echo = FALSE}
## build a reference table from the internal registry
## access the registry environment to get descriptions
registry_env = urbnindicators:::.table_registry
table_info = purrr::map_dfr(names(registry_env$tables), function(tn) {
  entry = registry_env$tables[[tn]]
  construct_names = if (!is.null(entry[["constructs"]])) {
    paste(purrr::map_chr(entry[["constructs"]], "name"), collapse = ", ")
  } else {
    tn
  }
  tibble::tibble(
    `Table name(s)` = construct_names,
    `ACS table(s)` = paste(entry[["acs_tables"]], collapse = ", "),
    Description = entry[["description"]])
})

reactable(
  table_info %>% dplyr::arrange(`Table name(s)`),
  columns = list(
    `Table name(s)` = colDef(minWidth = 130),
    `ACS table(s)` = colDef(minWidth = 80),
    Description = colDef(minWidth = 200)),
  filterable = TRUE)
```
