---
title: "The urbnindicators Codebook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{The urbnindicators Codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>")
```

```{r setup}
library(urbnindicators)
library(dplyr)
library(reactable)
```

Every dataframe returned by `compile_acs_data()` includes a codebook
as an attribute. The codebook documents every variable in the
dataset--what it represents, how it was calculated, and whether it is
a raw ACS estimate or a derived measure.

## Accessing the codebook

The codebook is stored as an attribute of the dataframe and can be
retrieved with `attr()`:

```{r}
df = compile_acs_data(years = 2024, geography = "us")
codebook = attr(df, "codebook")
```

## Understanding the columns

The codebook's three primary columns are:

-   **calculated_variable**: The variable name as it appears in the
    dataframe (e.g., `snap_received_percent`).

-   **variable_type**: The kind of variable. Values include `Count`
    (a raw ACS estimate directly from the API), `Percent` (a derived
    ratio), `Sum` (a derived count), `Median`, `Median ($)`,
    `Average`, `Quintile ($)`, `Index`, and `Metadata` (a
    non-computed field such as a geographic identifier).

-   **definition**: A description of how the variable was calculated.
    Raw ACS variables read `"This is a raw ACS estimate."`; derived
    variables spell out their inputs alongside the original Census
    Bureau variable codes, e.g.,
    `"Numerator = snap_received (B22003_002). Denominator = snap_universe (B22003_001)."`
    Subtracted components appear after a `-`.

The codebook also carries supporting columns used for margin-of-error
calculation and re-aggregation--list-columns naming each variable's
numerator/denominator components (`numerator_vars`,
`denominator_vars`, and their `_subtract_` counterparts), an
`se_calculation_type`, and an `aggregation_strategy` (how
`interpolate_acs()` treats the variable). Because margins of error
for derived variables are computed from these components, their
accuracy is critical.

## Browse the codebook

The interactive table below shows the three primary columns. Use the
search box to filter by variable name, type, or
definition text. Note that this codebook reflects all variables from
the tables returned by `list_tables()`, but if you were to specify
different tables in your `compile_acs_data()` call, your codebook
would comprise different variable listings. 

```{r, echo = FALSE}
reactable(
  codebook %>%
    transmute(
      Variable = calculated_variable,
      Type = variable_type,
      Definition = definition),
  columns = list(
    Variable = colDef(minWidth = 200),
    Type = colDef(minWidth = 50),
    Definition = colDef(minWidth = 100)),
  filterable = TRUE)
```
