---
title: "The urbnindicators Codebook"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{codebook}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  comment = "#>")
```

```{r setup}
library(urbnindicators)
library(dplyr)
library(reactable)
```

Every dataframe returned by `compile_acs_data()` includes a codebook
as an attribute. The codebook documents every variable in the
dataset--what it represents, how it was calculated, and whether it is
a raw ACS estimate or a derived measure.

## Accessing the codebook

The codebook is stored as an attribute of the dataframe and can be
retrieved with `attr()`:

```{r}
df = compile_acs_data(year = 2024, geography = "us")
codebook = attr(df, "codebook")
```

## Understanding the columns

The codebook has three columns:

-   **Variable**: The variable name as it appears in the dataframe
    (e.g., `snap_received_percent`).

-   **Type**: The kind of variable. Common types include:
    -   `count` -- a raw ACS estimate directly from the API
    -   `percent` -- a derived ratio, typically a count divided by a
        universe variable
    -   `metadata` -- a non-computed variable such as a median or
        geographic identifier

-   **Definition**: A formula describing how the variable was
    calculated. For raw ACS variables, this is the original Census
    Bureau variable code (e.g., `B22003_002`). For derived variables,
    this is an expression like `snap_received / snap_universe` that
    shows the numerator and denominator. More complex definitions may
    reference multiple raw variables joined with `+` in the numerator
    or denominator.

These definition strings are also used internally to calculate
margins of error for derived variables, so their accuracy is
critical.

## Browse the codebook

Use the search box below to filter by variable name, type, or
definition text. Note that this codebook reflects all variables from
the tables returned by `list_tables()`, but if you were to specify
different tables in your `compile_acs_data()` call, your codebook
would comprise different variable listings. 

```{r, echo = FALSE}
reactable(
  codebook %>%
    transmute(
      Variable = calculated_variable,
      Type = variable_type,
      Definition = definition),
  columns = list(
    Variable = colDef(minWidth = 200),
    Type = colDef(minWidth = 50),
    Definition = colDef(minWidth = 100)),
  filterable = TRUE)
```
