---
title: "coefficients-of-variation"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coefficients-of-variation}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(urbnindicators)
```

The Census Bureau API exposes an extensive set of estimates derived from responses to the American Community Survey. 
But it also exposes a companion set of margins of error (MOE). These MOEs are critical for accurately interpreting--
and responsibly using--ACS estimates. 

In brief, margins of error quantify the error associated with conducting a survey
and describe the precision of survey-derived statistics. If a point estimate is 500, and the associated 90% margin of error
is 100, then we would expect that if we were to use the same surveying methods to calculate that same estimate 100 times, 
90 of those 100 repeated survey efforts would produce point estimates in the range from 400 to 600. The remaining 10 would 
produce point estimates outside of that range (e.g., of 385 and 650).

This should give us some pause about our original point estimate; an MOE of 100 for a point estimate of 500 is fairly substantial.
However, all too often, analysts are unsure how to incorporate MOEs into their analysis, inference drawing, and data summary processes.
This can also make for more work as compared to just reporting out on point estimates. Further, for analysts working with derived measures--
e.g., those calculated by dividing two point estimates--it may be unclear how to evaluate the error associated with these newly-created measures.

`urbnindicators` helps to facilitate the incorporation of MOEs into analysts' workflows through a few inter-related functions. 
