---
title: "Aggregating to Custom Geographies"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{custom-geographies}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  dpi = 150,
  fig.width = 7,
  comment = "#>")
```

```{r setup, echo = FALSE}
library(dplyr)
library(ggplot2)
library(stringr)
library(urbnindicators)
library(sf)
library(urbnthemes)
library(tidycensus)
```

Census tracts are useful geographic units, but their small populations
often produce estimates with large margins of error. When analyzing
data, these imprecise estimates make it difficult to detect meaningful
differences between areas---even when real differences exist.

`calculate_custom_geographies()` addresses this by aggregating
tract-level data to user-defined geographies (e.g., neighborhoods,
planning districts, or school zones). This aggregation increases sample
sizes, reduces coefficients of variation, and enables more reliable
statistical inference.

# Example: DC Tract Data

We'll demonstrate using tract-level data for Washington, DC, comparing
the share of population receiving SNAP benefits across areas.

```{r, message = FALSE}
dc_tracts = compile_acs_data(
  years = 2024,
  geography = "tract",
  states = "DC",
  spatial = TRUE)
```

# Creating Custom Geographies

For illustration, we'll create four quadrants of DC by grouping tracts
based on their centroid coordinates. In practice, you would join tracts
to meaningful boundaries like neighborhoods or planning areas, or you could
group arbitrary numbers of adjacent tracts to form pseudo-neighborhoods with
larger populations than are captured in any single tract.

```{r}
# Calculate tract centroids and assign to quadrants
dc_tracts = dc_tracts %>%
  mutate(
    centroid = st_centroid(geometry),
    longitude = st_coordinates(centroid)[, 1],
    latitude = st_coordinates(centroid)[, 2],
    quadrant = case_when(
      longitude < median(longitude) & latitude >= median(latitude) ~ "Northwest",
      longitude >= median(longitude) & latitude >= median(latitude) ~ "Northeast",
      longitude < median(longitude) & latitude < median(latitude) ~ "Southwest",
      longitude >= median(longitude) & latitude < median(latitude) ~ "Southeast")) %>%
  select(-centroid, -longitude, -latitude)

# Aggregate to quadrants
dc_quadrants = calculate_custom_geographies(
  .data = dc_tracts,
  group_id = "quadrant",
  spatial = TRUE)
```

# Comparing Precision: Tracts vs. Custom Geographies

The maps below show the share of households receiving SNAP benefits.
Notice how aggregating to quadrants produces more precise estimates with
smaller coefficients of variation.

```{r, fig.height = 4}
# Tract-level map
map_tracts = dc_tracts %>%
  ggplot() +
  geom_sf(aes(fill = snap_received_percent), color = "white", linewidth = 0.1) +
  scale_fill_gradientn(
    colors = c("#CFE8F3", "#1696D2", "#0A4C6A"),
    limits = c(0, 0.5),
    labels = scales::percent,
    na.value = "grey80") +
  urbnthemes::theme_urbn_map() +
  theme(legend.position = "bottom") +
  labs(
    fill = "SNAP Receipt",
    title = "Tract-Level Estimates",
    subtitle = paste0("Median CV: ", round(median(dc_tracts$snap_received_percent_CV, na.rm = TRUE), 1)))

# Quadrant-level map
map_quadrants = dc_quadrants %>%
  ggplot() +
  geom_sf(aes(fill = snap_received_percent), color = "white", linewidth = 0.3) +
  scale_fill_gradientn(
    colors = c("#CFE8F3", "#1696D2", "#0A4C6A"),
    limits = c(0, 0.5),
    labels = scales::percent,
    na.value = "grey80") +
  urbnthemes::theme_urbn_map() +
  theme(legend.position = "bottom") +
  labs(
    fill = "SNAP Receipt",
    title = "Quadrant-Level Estimates",
    subtitle = paste0("Median CV: ", round(median(dc_quadrants$snap_received_percent_CV, na.rm = TRUE), 1)))

map_tracts
map_quadrants
```

The quadrant-level estimates have substantially lower CVs, indicating
more reliable estimates. This precision gain enables meaningful
statistical comparisons.

# Detecting Statistically Significant Differences

With more precise estimates, we can better identify where SNAP receipt
rates differ significantly from the citywide average.

```{r, fig.height = 4}
# Calculate DC-wide SNAP rate for comparison
dc_snap_rate = sum(dc_tracts$snap_received, na.rm = TRUE) /
               sum(dc_tracts$snap_universe, na.rm = TRUE)

# Test significance at tract level
tracts_sig = dc_tracts %>%
  mutate(
    significant = tidycensus::significance(
      est1 = snap_received_percent,
      est2 = dc_snap_rate,
      moe1 = snap_received_percent_M,
      moe2 = 0.005,
      clevel = 0.9),
    difference = case_when(
      !significant ~ "Not significant",
      snap_received_percent > dc_snap_rate ~ "Higher than DC average",
      snap_received_percent < dc_snap_rate ~ "Lower than DC average"))

# Test significance at quadrant level
quadrants_sig = dc_quadrants %>%
  mutate(
    significant = tidycensus::significance(
      est1 = snap_received_percent,
      est2 = dc_snap_rate,
      moe1 = snap_received_percent_M,
      moe2 = 0.005,
      clevel = 0.9),
    difference = case_when(
      !significant ~ "Not significant",
      snap_received_percent > dc_snap_rate ~ "Higher than DC average",
      snap_received_percent < dc_snap_rate ~ "Lower than DC average"))

# Color palette
sig_colors = c(
  "Higher than DC average" = "#1696D2",
  "Lower than DC average" = "#FDBF11",
  "Not significant" = "#D2D2D2")

# Maps
map_tracts_sig = tracts_sig %>%
  ggplot() +
  geom_sf(aes(fill = difference), color = "white", linewidth = 0.1) +
  scale_fill_manual(values = sig_colors, na.value = "grey80") +
  urbnthemes::theme_urbn_map() +
  theme(legend.position = "bottom") +
  labs(fill = "", title = "Tract-Level")

map_quadrants_sig = quadrants_sig %>%
  ggplot() +
  geom_sf(aes(fill = difference), color = "white", linewidth = 0.3) +
  scale_fill_manual(values = sig_colors, na.value = "grey80") +
  urbnthemes::theme_urbn_map() +
  theme(legend.position = "bottom") +
  labs(fill = "", title = "Quadrant-Level")

gridExtra::grid.arrange(
  map_tracts_sig, map_quadrants_sig,
  ncol = 2,
  top = grid::textGrob(
    "Aggregation Enables Detection of Statistically Significant Differences",
    gp = grid::gpar(fontsize = 12, fontface = "bold")))
```

At the tract level, large margins of error prevent many estimates from
reaching statistical significance. At the quadrant level, the same
underlying data produces estimates precise enough to detect significant
differences from the citywide average.

# Key Takeaways

1.  **Aggregation improves precision**: Combining tracts into larger
    geographies reduces CVs and margins of error.

2.  **Better inference**: More precise estimates enable detection of
    statistically significant differences that would otherwise be
    obscured by sampling error.

3.  **Proper error handling**: `calculate_custom_geographies()` correctly
    recalculates MOEs, SEs, and CVs for aggregated estimates following
    Census Bureau methodology---simply averaging tract-level statistics
    would produce incorrect results.

4.  **Flexible boundaries**: The function works with any custom geography
    defined by a grouping variable, allowing analysis at neighborhood,
    district, or other policy-relevant scales.
