---
title: "Custom Derived Variables"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Custom Derived Variables}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE)
```

```{r setup}
library(dplyr)
library(tidyr)
library(stringr)
library(urbnindicators)
library(tidycensus)
```

`compile_acs_data()` supports a large set of pre-built derived
variables (percentages, complements, etc.) from the package's
registered tables. But you can also define your own derived
variables using the `define_*()` helpers and pass them directly to
`compile_acs_data()` via the `tables` parameter. Your custom
variables automatically get codebook entries and margins of error,
just like the built-in ones.

This vignette walks through the process of identifying variables,
choosing denominators, and writing definitions, using household
type data (ACS table B11001) as a running example.

## The `define_*()` Helpers

| Helper | Use case | Key arguments |
|---|---|---|
| `define_percent()` | Single output: numerator / denominator | `output`, `numerator`, `denominator` (simple) or `numerator_variables`, `denominator_variables`, regex variants (complex) |
| `define_across_percent()` | Percentages across a regex-matched set of columns | `input_regex`, `output_suffix`, `denominator` or `denominator_function`, optional `exclude_regex` |
| `define_across_sum()` | Summing paired columns (e.g., male + female) | `input_regex`, `addend_function`, `output_naming_function` |
| `define_one_minus()` | Complement of an existing percentage (1 - x) | `output`, `source_variable` |
| `define_metadata()` | Non-computed variables (e.g., medians) | `output`, `definition_text` |

Definitions are passed to `compile_acs_data()` alongside table
names in a `list()`. They operate on the columns produced by the
tables you request, so make sure to include the table(s) whose
variables your definitions reference.

## Step 1: Identify the Raw Variables

We're going to create a variable describing the share of white
men between the ages of 18 and 44. The needed raw variables are
from ACS table B01001A, as described in the codebook

```{r}
codebook = get_acs_codebook()

codebook %>%
  filter(table == "B01001A") %>%
  pull(variable_clean) %>%
  head()
```

## Step 2: Choose Denominators

Selecting an appropriate denominator for percentage variables is
critical, and at times, complex. The basic approach is to simply
divide every variable by the table universe, which should contain
`universe` in the variable name. But in other cases, alternate
variables may make for more insightful denominators. For simplicity,
we just use the table universe here.

## Step 3: Write and Use Definitions

We express the new variable(s) using `define_*()` helpers and pass
the definitions directly to `compile_acs_data()`.

Our target variable--the share of white men aged 18-44--requires
summing several age-group columns (18-19, 20, 21, 22-24, 25-29,
30-34, 35-39, 40-44) and dividing by the table universe.
`define_percent()` with `numerator_regex` handles this: it matches
all columns fitting a pattern and sums them into the numerator.

The auto-processed column names from B01001A follow the pattern
`sex_by_age_white_alone_male_{age_range}_years`, so we can target
the 18-44 age groups with a regex that matches the leading digits
of each range:

```{r}
df = compile_acs_data(
  tables = list(
    "B01001A",
    define_percent(
      "white_male_18_44_percent",
      numerator_regex = "sex_by_age_white_alone_male_(1[89]|2[0-9]|3[0-9]|4[0-4])",
      numerator_exclude_regex = "percent",
      denominator_variables = c("sex_by_age_white_alone_universe"))),
  years = 2024,
  geography = "county",
  states = "NJ")

df %>%
  select(GEOID, white_male_18_44_percent) %>%
  head()
```

The definition is executed after the raw ACS data are fetched and
renamed, and the results appear in the codebook with automatically
calculated margins of error:

```{r}
attr(df, "codebook") %>%
  filter(calculated_variable == "white_male_18_44_percent") %>%
  pull(definition)
```

## More Examples

### Simple percentage

One numerator, one denominator:

```{r}
df_snap = compile_acs_data(
  tables = list(
    "snap",
    define_percent(
      "snap_received_percent_manual",
      numerator = "snap_received",
      denominator = "snap_universe")),
  years = 2024,
  geography = "county",
  states = "NJ")

## verify it matches the built-in version
df_snap %>%
  transmute(
    builtin = snap_received_percent,
    manual = snap_received_percent_manual,
    match = abs(builtin - manual) < 1e-10) %>%
  filter(!match)
```

### Across-percent with a complement

Percentages for all race categories, plus a person-of-color
complement. This is how the built-in `race` table works internally.

```r
compile_acs_data(
  tables = list(
    "race",
    define_across_percent(
      input_regex = "^race_nonhispanic|^race_hispanic",
      output_suffix = "_percent",
      denominator = "race_universe"),
    ## race_personofcolor_percent is the share of all individuals who are not
    ## non-Hispanic, White alone, i.e., the complement
    define_one_minus(
      "race_personofcolor_percent",
      source_variable = "race_nonhispanic_white_alone_percent")),
  years = 2024,
  geography = "county",
  states = "DC")
```

### Across-sum followed by across-percent

Sum male + female counts into combined age variables, then calculate
percentages. This is how the built-in `sex_by_age` table works
internally.

This one is a bit tricky--the source table includes variables for
each age group, split by sex. To get age groups, we have to add the
two sex-specific estimates for the given age group. This requires us
to specify an `input_regex`, which selects, in this case, all
female-specific age variables. The `addend_function` then
programmatically identifies the same-named, male-specific variables.
The `output_naming_function` simplifies the resulting combined
variable, removing the sex category and other extraneous words.

```r
compile_acs_data(
  tables = list(
    "sex_by_age",
    define_across_sum(
      input_regex = "sex_by_age_female_.*years($|_over$)",
      addend_function = function(column) {
        column %>% stringr::str_replace("female", "male")},
      output_naming_function = function(column) {
        column %>% stringr::str_replace("sex_by_age_female_", "age_")
      }),
    define_across_percent(
      input_regex = "^age.*years($|_over$)",
      output_suffix = "_percent",
      denominator = "sex_by_age_universe")),
  years = 2024,
  geography = "county",
  states = "DC")
```

### Complex percentage with subtraction

For cases where the numerator or denominator requires summing or
subtracting multiple variables, use the `numerator_variables`,
`denominator_variables`, and optional `*_subtract_*` arguments:

```{r}
df_complex = compile_acs_data(
  tables = list(
    "snap",
    define_percent(
      "snap_not_received_pct_complex",
      numerator_variables = c("snap_universe"),
      numerator_subtract_variables = c("snap_received"),
      denominator_variables = c("snap_universe"))),
  years = 2024,
  geography = "county",
  states = "DC")

df_complex %>%
  mutate(sums_to_one = snap_received_percent + snap_not_received_pct_complex) %>%
  glimpse()
```

## Verify Results

A few strategies for quality-checking custom derived variables:

1.  **Compare to a published benchmark.** Percentages reported by the
    Census Bureau in the Subject Tables (tables prefixed with `S`)
    can serve as reference values for derived variables computed from
    the detailed tables (prefixed with `B` or `C`).

2.  **Manually compute a benchmark.** Identify the relevant numerator
    and denominator variables and manually calculate the derived
    variable, then compare. This is especially useful for complex
    definitions where the numerator is itself a sum.

3.  **Plot histograms.** Use `pivot_longer()` and `facet_wrap()` to
    check for unexpected spikes or outlier values across a series of
    related variables.

4.  **Check for missingness.** Derived variables should generally have
    low or no missingness; substantial missingness may indicate a
    calculation error.

5.  **Inspect the codebook.** Verify that `attr(df, "codebook")`
    accurately documents your custom variable's definition. The
    definition string drives the margin of error calculation, so
    errors there will propagate to MOEs.
