---
title: "How to Open a Pull Request"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Open a Pull Request}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(dplyr)
library(tidyr)
library(stringr)
library(urbnindicators)
library(tidycensus)
```

Pull requests (PRs) are a way to propose changes to the codebase of a
project. They allow you to suggest changes, improvements, or fixes to
the code, and they provide a platform for discussion and review before
those changes are merged into the main codebase.

For `library(urbnindicators)`, a common use-case for a PR is to propose
a new variable or series of variable. This vignette will illustrate the
complete PR process for a new series of variables, including where the
PR should propose codebase updates and the supporting tests and
verifications submitted with the PR to assert that the newly-created
variables are correctly calculated and appropriately documented.

## Overview

There are five places in the codebase that should either be updated or
quality-checked as part of the process of adding a new variable(s).
These include:

1.  `R/list_acs_variables.R`: this is where the desired variable codes
    are selected before being passed to `tidycensus::get_acs()`.

2.  `R/compile_acs_data.R`, specifically within
    `internal_compute_acs_variables()`: this is where new derived
    variables are created, e.g., by standardizing counts by their table
    denominators.

3.  `R/generate_codebook.R`: this is where we use meta-programming to
    evaluate the code in `R/compile_acs_data.R` to document the
    variables and operations used to calculate derived variables.

4.  `R/calculate_cvs.R`: this is where we leverage metadata from the
    codebook to calculate coefficients of variation (and other measures
    of error) for all of the returned variables.

5.  `R/make_pretty_names.R`: this facilitates using variable names in
    charts and tables, so we want to ensure that the
    programmatically-generated names of new variables return reasonable
    pretty-printed names.

Our last step is to integrate our proposed changes, execute
`compile_acs_data()`, and evaluate the results to ensure that the
changes not only work in isolation, but that they also work as part of
the broader codebase.

## Our Variable Series

We're going to add a series of estimates that describe "Household Type",
which are contained in table B11001. We can find the relevant variables
by navigating the codebook returned by `tidycensus::load_variables()`.

```{r}
codebook = load_variables(dataset = "acs5", year = 2023)

codebook %>%
  filter(str_detect(name, "B11001")) %>%
  head()
```

## 1: `R/list_acs_variables()`

We can see how these variables will be renamed behind the scenes in
`library(urbnindicators)` by providing the table name to
`urbnindicators::select_variables_by_name()`.

These are auto-named reasonably well, but we can make the names more
concise by removing the repeated substring `"including_living_alone_"`.
Note that each variable names ends in an `_`; this is intentional.

This code is then the code that should be added within
`list_acs_variables()` under the appropriate commented header, which in
this case is `####----HOUSEHOLD COMPOSITION----####`. We'll then add the
table name in comments above this new code, so that the resulting block
of code will look like the below:

```         
   ####----EMPLOYMENT----####
      ## EMPLOYMENT STATUS FOR THE POPULATION 16 YEARS AND OVER
      employment_civilian_labor_force_universe_ = "B23025_003",
      employment_civilian_labor_force_employed_ = "B23025_004",

   ####----HOUSEHOLD COMPOSITION----####
      ## AVERAGE HOUSEHOLD SIZE OF OCCUPIED HOUSING UNITS BY TENURE
      household_size_average_ = "B25010_001",
      household_size_average_owneroccupied_ = "B25010_002",
      household_size_average_renteroccupied_ = "B25010_003",
      
      ## HOUSEHOLD TYPE (INCLUDING LIVING ALONE)
      select_variables_by_name("B11001_", census_codebook = codebook) %>%
        stats::setNames(names(.) %>% stringr::str_remove_all("including_living_alone_")),
   
   ####----HEALTH INSURANCE----####
      ## HEALTH INSURANCE COVERAGE STATUS AND TYPE BY EMPLOYMENT STATUS
      select_variables("B27011"),
```

## 2: `R/compile_acs_data`

Now we need to calculate any desired variables. In virtually all cases
(excepting, for example, estimates that reflect non-count quantities,
such as medians or percentiles), we will want to calculate percentages.
In some cases, however, we will first want to create intermediate count
variables, such as by summing or subtracting multiple count variables.
In this case, we'll just calculate percentages.

Selecting an appropriate denominator for percentage variables is
critical, and at times, complex. The basic approach is to simply divide
every variable by the table universe, which should contain `universe` in
the variable name. But in other cases, alternate variables may make for
more insightful denominators. For example, if we were interested in the
share of family households that were headed by single male householders,
we would want to divide
`household_type_family_households_other_family_male_householder_no_spouse_present`
by `household_type_family_households_` rather than by
`household_type_universe_`.

At this point, we'll want to obtain some sample data to ensure our
first-pass code is functional. We can do this as so:

```{r}
sample_data = tidycensus::get_acs(
  years = 2022,
  geography = "county",
  state = "NJ",
  variables = select_variables_by_name("B11001_", census_codebook = codebook) %>%
        stats::setNames(names(.) %>% stringr::str_remove_all("including_living_alone_")),
  output = "wide") %>% ## urbnindicators returns wide, not long, data
  ## when urbnindicators calculates variables, it does so after de-selecting all 
  ## MOE variables (it joins them back later in the process)
  select(GEOID, NAME, matches("_E")) %>% 
  ## urbnindicators also systematically drops all "_E" suffices, which denote "ESTIMATE"
  ## variables
  rename_with(cols = everything(), ~ str_remove_all(.x, "_E")) 

sample_data %>%
  head()
```

We calculate percentages across a series of variables using
`dplyr::across()`; the general pattern is:

```         
dplyr::across(
  .cols = dplyr::matches("[UNIQUE REGEX TO MATCH DESIRED NUMERATOR VARIABLES]),
  .fns = ~ .x / [UNIVERSE VARIABLE],
  .names = "{.col}_percent")
```

In our case, that will be as follows. Note that we explicitly exclude
the universe variable from our selection of numerators--this is an easy
step to forget!

Defining an appropriate selection term for the numerators can be
particularly challenging because we must ensure both that we select the
desired variables and that we don't unintentionally select any undesired
variables, which can be especially difficult when applied to the real
case, where we're not selecting from our `sample_data` object but from
the full suite of variables returned by `urbnindicators`. To help
evaluate and prevent such issues, users should call `compile_acs_data()`
and ensure that the selection term they develop during this stage does
not select any of the variables that are already returned by
`compile_acs_data()`.

```{r}
sample_data %>%
  dplyr::transmute(
    dplyr::across(
      .cols = c(dplyr::matches("household_type"), -dplyr::matches("universe")),
      .fns = ~ .x / household_type_universe,
      .names = "{.col}_percent")) %>%
  ## just to print out a subset of our results
  head() %>% select(1:3)
```

If we're satisfied with the resulting estimates, we can add this code
into `compile_acs_data.R`, within `internal_compute_acs_variables()`.
Akin to our preceding step, this new code should be organized under the
`####----HOUSEHOLD COMPOSITION----####` comment header.

## 3: `R/generate_codebook.R`

If all has gone well, there will be no new code to add to
`generate_codebook()`-- new variables should be automatically documented
by the function and included in the resulting codebook object that is
attached as an attribute to the primary dataframe returned by
`compile_acs_data()`. However, we should meticulously verify that each
new variable is in fact documented in the codebook, and that its
documentation is accurate. If the documentation is incorrect, the error
we estimate for derived variables will also be incorrect.

Note that if the documentation is incorrect, the user creating the PR
should still open the PR and should note the issue with the
documentation; the PR should not include any changes to
`generate_codebook()`.

## 4: `R/calculate_cvs.R`

Like with step 3 above, if all has gone well, there will be no new code
to add to `calculate_cvs()`. Here again, the primary task is to check
that coefficients of variation appear reasonable (and if not, to flag
any concerns in the PR).

At this stage, users should also check the magnitude of errors for all
variables --raw ACS estimates and `urbnindicators`-calculated variables
alike--across smaller geographies, such as all tracts in one or more
states. If CVs are large (e.g., over 50) for a large share of all
tracts, this may indicate that the series of interest is not appropriate
for tract-level analysis. Because `urbnindicators` is designed to
facilitate tract-level analysis, variables that are consistently
unreliable at the tract-level will not be integrated into the codebase.

## 5: `R/make_pretty_names.R`

Finally, we need to ensure that the new variables have reasonable names.
We can evaluate this easily with our sample data:

```{r}
sample_data %>%
  dplyr::mutate(
    dplyr::across(
      .cols = c(dplyr::matches("household_type"), -dplyr::matches("universe")),
      .fns = ~ .x / household_type_universe,
      .names = "{.col}_percent")) %>%
  urbnindicators::make_pretty_names() %>%
  colnames()
```

Though these names are far from perfect, they're reasonably concise and
descriptive, and there are no problematic words that are lost in the
pretty-ifying process (e.g., acronyms, series of numbers, etc.). We can
leave it to users to make other adjustments, such as removing the
substring "Household Type ", if they want even more concise names.

## 6: Evaluate Changes by Calling `compile_acs_data()`

Lastly, users should ensure that the code they've developed in isolation
for their PR functions as intended, and without unanticipated side
effects, by integrating their proposed changes into the codebase (on
their branch), loading the current version of `urbnindicators` (which is
done by calling `devtools::load_all()` in the console), and then calling
`compile_acs_data()` in a scratch file and interactively exploring both
the data and codebook.

## 7: Open the PR

Once users are satisfied with the proposed code (and/or have noted any
issues), they should click on their branch in the GitHub repository and
click `New pull request`.
