---
title: "How to Open a Pull Request"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{How to Open a Pull Request}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  warning = FALSE,
  message = FALSE)
```

```{r setup}
library(dplyr)
library(tidyr)
library(stringr)
library(urbnindicators)
library(tidycensus)
```

Pull requests (PRs) are a way to propose changes to the codebase of a
project. They allow you to suggest changes, improvements, or fixes to
the code, and they provide a platform for discussion and review before
those changes are merged into the main codebase.

For `library(urbnindicators)`, a common use-case for a PR is to propose
a new variable or series of variables. This vignette will illustrate
the complete PR process for a new series of variables. The core of the
PR is a single `register_table()` call in `R/table_registry.R` that
defines the raw ACS variables and specifies how derived variables are 
computed. The codebook and measures of error are all generated automatically 
from the table registry.

## Overview

Adding a new table requires a single code change:

1.  **`R/table_registry.R`**: Add a `register_table()` call that
    declares (a) the raw ACS variable codes and (b) a `definitions`
    list that uses the `define_*` helpers to specify how derived
    variables (e.g., percentages) are calculated.

After writing the registration, the PR should
include quality checks to verify that the new variables are correctly
calculated and appropriately documented.

The quality-check steps are:

2.  **Codebook**: Verify that each new variable is documented in the
    codebook and that its documentation is accurate.
3.  **Coefficients of variation**: Verify that CVs appear reasonable.
4.  **Pretty names**: Verify that `make_pretty_names()` consistently 
    clearly renames the new variables.
5.  **Integration test**: Call `compile_acs_data()` end-to-end and
    inspect the results.

## Our Variable Series

We're going to add a series of estimates that describe "Household
Type (White Alone)", which are contained in table B11001. We can find the relevant
variables by navigating the codebook returned by
`tidycensus::load_variables()`.

```{r}
codebook = load_variables(dataset = "acs5", year = 2024)

codebook %>%
  filter(str_detect(name, "B11001")) %>%
  head(2) %>%
  glimpse()
```

## Step 1: Identify the Raw Variables

We can see how these variables will be renamed behind the scenes in
`library(urbnindicators)` by providing the table name to
`urbnindicators::select_variables_by_name()`.

```{r}
select_variables_by_name("B11001_", census_codebook = codebook) %>%
  stats::setNames(names(.) %>% str_remove_all("including_living_alone_"))
```

These are auto-named reasonably well, but we can make the names more
concise by removing the repeated substring `"including_living_alone_"`.
Note that each variable name ends in an `_`; this is intentional.

## Step 2: Choose Denominators for Derived Variables

Selecting an appropriate denominator for percentage variables is
critical, and at times, complex. The basic approach is to simply divide
every variable by the table universe, which should contain `universe`
in the variable name. But in other cases, alternate variables may make
for more insightful denominators. For example, if we were interested
in the share of family households that were headed by single male
householders, we would want to divide
`household_type_family_households_other_family_male_householder_no_spouse_present`
by `household_type_family_households_` rather than by
`household_type_universe_`.

At this point, we'll want to obtain some sample data to verify that our
variable choices and denominator logic are correct:

```{r}
sample_data = tidycensus::get_acs(
  years = 2024,
  geography = "county",
  state = "NJ",
  variables = select_variables_by_name("B11001_", census_codebook = codebook) %>%
    stats::setNames(names(.) %>% stringr::str_remove_all("including_living_alone_")),
  output = "wide") %>%
  ## selecting only estimate ("_E") variables, not margin of error ("_M") variables
  select(GEOID, NAME, matches("_E")) %>%
  rename_with(cols = everything(), ~ str_remove_all(.x, "_E"))

sample_data %>%
  head(2) %>%
  glimpse()
```

We can test our percentage calculations on the sample data. Note the
use of `urbnindicators::safe_divide()`, which returns `0` rather than
`NaN` when the denominator is zero.

```{r}
sample_data %>%
  transmute(
    across(
      .cols = c(matches("household_type"), -matches("universe")),
      .fns = ~ safe_divide(.x, household_type_universe),
      .names = "{.col}_percent")) %>%
  head() %>% 
  glimpse()
```

## Step 3: Write the `register_table()` Call

Once we're satisfied with the raw variables and derived logic, we
express it as a `register_table()` call. This is the primary code
change in the PR.

The `definitions` list uses the `define_*` helpers to declaratively
specify how each derived variable is computed. The package uses this
specification to both execute the computation and auto-generate the
codebook documentation and error calculations.

### The `define_*` Helpers

| Helper | Use case | Key arguments |
|---|---|---|
| `define_percent()` | Single output: numerator / denominator | `output`, `numerator`, `denominator` (simple) or `numerator_variables`, `denominator_variables`, regex variants (complex) |
| `define_across_percent()` | Percentages across a regex-matched set of columns | `input_regex`, `output_suffix`, `denominator` or `denominator_function`, optional `exclude_regex` |
| `define_across_sum()` | Summing paired columns (e.g., male + female) | `input_regex`, `addend_function`, `output_naming_function` |
| `define_one_minus()` | Complement of an existing percentage (1 - x) | `output`, `source_variable` |
| `define_metadata()` | Non-computed variables (e.g., medians) | `output`, `definition_text` |

### Household type registration

For our household type example, the `register_table()` call is:

```r
register_table(list(
  name = "household_type",
  description = "Household type (including living alone)",
  acs_tables = "B11001",
  depends_on = character(0),
  raw_variable_source = list(type = "select_variables"),
  raw_variables_transform = function(vars) {
    stats::setNames(vars, names(vars) %>%
      stringr::str_remove_all("including_living_alone_"))
  },
  definitions = list(
    define_across_percent(
      input_regex = "^household_type",
      exclude_regex = "universe|percent",
      output_suffix = "_percent",
      denominator = "household_type_universe"))
))
```

### More examples

**Simple percentage** (one numerator, one denominator):

```r
## from the SNAP table
definitions = list(
  define_percent("snap_received_percent",
                 numerator = "snap_received",
                 denominator = "snap_universe"))
```

**Across-percent with a complement** (percentages for all race
categories, plus a person-of-color complement):

```r
## from the race table
definitions = list(
  define_across_percent(
    input_regex = "^race_nonhispanic|^race_hispanic",
    exclude_regex = NULL,
    output_suffix = "_percent",
    denominator = "race_universe"),
  ## race_personofcolor_percent is the share of all individuals who are not
  ## non-Hispanic, White alone, i.e, the complement
  define_one_minus("race_personofcolor_percent",
                   source_variable = "race_nonhispanic_white_alone_percent"))
```

**Across-sum followed by across-percent** (sum male + female counts
into combined age variables, then calculate percentages):

This one is a bit tricky--the source table includes variables for each
age group, split by sex. To get age groups, we have to add the two
sex-specific estimates for the given age group. This requires us to specify
an `input_regex`, which selects, in this case, all female-specific age variables.
The `addend_function` then programmatically identifies the same-named, male-specific 
variables. The `output_naming_function` simplifies the resulting combined variable,
removing the sex category and other extraneous words.

```r
## from the sex_by_age table
definitions = list(
  define_across_sum(
    input_regex = "sex_by_age_female_.*years($|_over$)",
    addend_function = function(column) {
      column %>% stringr::str_replace("female", "male")
    },
    output_naming_function = function(column) {
      column %>% stringr::str_replace("sex_by_age_female_", "age_")
    }),
  define_across_percent(
    input_regex = "^age.*years($|_over$)",
    output_suffix = "_percent",
    denominator = "sex_by_age_universe"))
```

### Placement in the file

The `register_table()` call should be added to `R/table_registry.R`
under the appropriate comment header. For our example, this would be
under `####----TABLE REGISTRATIONS: HOUSEHOLD COMPOSITION----####`.

## Step 4: Verify the Codebook

After writing the registration, load the package with
`devtools::load_all()` and call `compile_acs_data()`. New variables should be
automatically documented by `generate_codebook()` and included in the
codebook. Verify that each new variable is documented and that its documentation
is accurate. If the documentation is incorrect, the error we estimate
for derived variables will also be incorrect.

If the documentation is incorrect, the PR should note the issue.

## Step 5: Verify Coefficients of Variation

Like the codebook, CVs are computed automatically from the
`definitions` metadata. Check that coefficients of variation appear
reasonable.

Users should also check the magnitude of errors for all variables--raw
ACS estimates and derived variables alike--across
smaller geographies, such as all tracts in one or more states. If CVs
are large (e.g., over 50) for a large share of all tracts, this may
indicate that the series of interest is not appropriate for
tract-level analysis. Because `urbnindicators` is designed to
facilitate tract-level analysis, tables that are consistently
unreliable at the tract level will not be integrated into the codebase.

## Step 6: Verify Pretty Names

Ensure that the new variables have reasonable names:

```{r}
sample_data %>%
  mutate(
    across(
      .cols = c(matches("household_type"), -matches("universe")),
      .fns = ~ .x / household_type_universe,
      .names = "{.col}_percent")) %>%
  make_pretty_names() %>%
  colnames()
```

Though these names are far from perfect, they're reasonably concise
and descriptive, and there are no problematic words that are lost in
the pretty-ifying process (e.g., acronyms, series of numbers, etc.).
We can leave it to users to make other adjustments, such as removing
the substring "Household Type ", if they want even more concise names.

## Step 7: Quality Check Results

There are a few strategies for quality-checking the results of a
series of variables:

1.  Compare to a published benchmark value. This works well for
    derived variables, such as percentages, that are reported by the
    Census Bureau in the Subject Tables (tables prefixed with an `S`)
    but that are not directly available via the detailed tables
    (prefixed with a `B` or `C`). However, for this series of
    variables, there are not counterpart percentages reported as part
    of the subject tables. (An example of one of multiple reasons that
    `urbnindicators` exclusively uses data from the detailed tables.)

2.  Manually compute a benchmark value.
    Identify the relevant numerator and denominator variables
    (in the case of a derived percentage) and manually calculate the
    derived variable, then compare the manually-computed benchmark to
    the programmatically-calculated version. This seems very simple
    for our example here, where each derived variable is a given
    variable divided by the table universe, but with more complex
    variables--e.g., where a numerator is a summed variable
    itself--this is a very useful quality check.

3.  Plot a histogram of the computed variable(s). If there are multiple
    variables in a series, use `pivot_longer()` to turn the dataframe
    long, then use `facet_wrap()` to plot each histogram side-by-side.
    Check for unexpected spikes and outlier values.

4.  Check for missingness. Generally, derived variables should have
    low or no missingness, so any substantial number of missing
    observations may be an indication that a calculation has gone
    awry.

## Step 8: Open the PR

Once users are satisfied with the proposed code (and/or have noted any
issues), they should click on their branch in the GitHub repository
and click `New pull request`.
