---
title: "Drawing Inferences while Accounting for Survey Error"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Drawing Inferences while Accounting for Survey Error}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options:
  markdown:
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  dpi = 600,
  fig.width = 7,
  comment = "#>")
```

```{r setup, echo = FALSE}
library(dplyr)
library(ggplot2)
library(tidyr)
library(stringr)
library(urbnindicators)
library(sf)
library(urbnthemes)
library(tidycensus)
library(gridExtra)
```

Because data from the American Community **Survey** are--you guessed
it--survey data, they are subject to sampling error. Sampling error
means, in brief, that when we extrapolate from our sample--say, 200
households out of a neighborhood population of 1,000 households--we make
assumptions that the sample is representative of the population.[^1]
These assumptions are inherently imperfect, and so the estimates we use
based on sample responses have a quantifiable amount of error associated
with them.

[^1]: In reality, the "assumptions" used by the Census Bureau to
    extrapolate from survey responses to population estimates are very
    sophisticated. When working with individual-level data via the ACS
    Public Use Microdata Sample (PUMS), the error generated by these
    assumptions can be calculated by using a set of 80 variables that
    describe "replicate weights". But when working with pre-tabulated
    ACS estimates, as is the case here, the error generated by these
    assumptions is conveniently reported alongside the associated
    estimate as margin of error.

Using the known error around an estimate allows us to better say what we
do and do not know about the population. Without this information, we
might say, based on our sample of 200 households, that 5% of
neighborhood households are housing cost-burdened renter households. But
when we factor in the error around the estimate, we might say something
more nuanced, like: "Between 3.5% and 6.5% of households are housing
cost-burdened renter households." This reflects that survey estimates
are more accurately thought of as a range of likely values rather than a
single absolute value.

Using quantified error around survey estimates will allow us to do three
(at least) critical classes of activities to improve our
inference-making with these data:

1.  **Evaluate estimate quality** to inform whether we should, for
    example, aggregate our data to create more precise estimates or
    select a different measure or data source.

2.  **Conduct statistical significance testing** to evaluate whether an
    estimate is statistically significantly greater than another
    estimate(s).

3.  **Report and visualize error** to our audiences so that they can
    understand the limitations of the data.

Below, we illustrate how `urbnindicators` facilitates each of these
three classes of activities. But first, a very quick overview of
measures of error.

-   **Margins Of Error (MOE)** are our baseline measurement of error.
    The ACS reports MOEs at a 90% confidence level, which enables us to
    say something like: "20% (±5%) of households in the neighborhood
    have no access to a car." (And then we should include, either as a
    footnote or in the body of the document, that this and other MOEs
    are calculated at the 90% confidence level). What this means in
    practice is that if we were to repeat 100 times–using exactly the
    same methods–our approach to calculating this estimate, 90 of those
    times we would produce a parallel estimate between 15% and 25%,
    while 10 of those times, our estimate would fall outside this range.

-   **Standard Errors (SE)** are derived from MOEs by dividing the MOE
    against a confidence level-related value. 90% SEs are calculated by
    dividing an MOE by 1.645.

-   **Coefficients of Variation (CV)** relate error to the size of the
    estimate. They are calculated by dividing the SE by the estimate and
    then multiplying by 100. CVs are helpful because they provide a
    unit-agnostic measurement of variable quality: a CV of 40 for a
    given variable means the same thing as a CV of 40 for another
    unrelated variable, whereas MOEs and SEs are not directly comparable
    across variables.

# Evaluate Estimate Quality

MOEs allow us to assess whether estimates have problematically large
errors. While there's not a right-or-wrong threshold for what
constitutes a problematic MOE, a common approach is to calculate the
coefficient of variation (CV = SE / estimate * 100, where SE = MOE /
1.645) and use thresholds between 30 and 40.

As shown below, variables that rely on larger sample sizes tend to have
smaller MOEs. Typically, there are two strategies to reduce error: (1)
aggregate estimates, either across geographies or across variables, or
(2) use larger geographies.

```{r, warning = FALSE, message = FALSE}
acs_df_county = compile_acs_data(
  years = c(2024),
  tables = "age",
  geography = "county",
  states = "NJ")

acs_df_tract = compile_acs_data(
  years = c(2024),
  tables = "age",
  geography = "tract",
  states = "NJ",
  spatial = TRUE)
```

```{r, message = FALSE, warning = FALSE, fig.width = 7, fig.height = 12}
## derive CVs from MOEs for quality evaluation
plot_df = bind_rows(
      acs_df_county %>% mutate(geography = "County"),
      acs_df_tract %>% mutate(geography = "Tract")) %>%
    st_drop_geometry() %>%
    select(
      geography,
      c(matches("^age.*percent$") & matches("(_6|_7|_8)")),
      c(matches("^age.*percent_M$") & matches("(_6|_7|_8)"))) %>%
    pivot_longer(-geography) %>%
    mutate(
      base_var = str_remove(name, "_M$"),
      type = if_else(str_detect(name, "_M$"), "moe", "estimate")) %>%
    pivot_wider(
      id_cols = c(geography, base_var),
      names_from = type,
      values_from = value,
      values_fn = list) %>%
    tidyr::unnest(c(estimate, moe)) %>%
    mutate(
      cv = (moe / 1.645) / estimate * 100,
      cv = if_else(is.infinite(cv), NA_real_, cv),
      name = base_var %>%
        str_replace_all(c("_" = " ", "percent" = "(%)")) %>%
        str_remove_all("age") %>%
        str_squish() %>% str_trim(),
      plot_title = str_c(geography, ": ", name))

factor_levels = plot_df %>%
  arrange(name) %>%
  distinct(plot_title, .keep_all = TRUE) %>%
  pull(plot_title)

plot_df %>%
  mutate(
    plot_title = factor(plot_title, levels = factor_levels, ordered = TRUE)) %>%
  ggplot() +
    geom_histogram(
      aes(x = cv), fill = palette_urbn_main[1], bins = 50) +
    geom_vline(xintercept = 30, linetype = "dashed") +
    facet_wrap(~ plot_title, ncol = 2, scales = "free") +
    scale_x_continuous(limits = c(0, 150)) +
    guides(fill =  "none") +
    theme_urbn_print() +
    theme(axis.text.y = element_blank()) +
    labs(
      x = "CV (derived from MOE)",
      y = "Distribution",
      title = "County CVs are Smaller than Tract CVs\nCVs for Wider Age Ranges are Smaller than CVs for Narrow Age Ranges",
      subtitle = "Coefficients of variation at the county and tract levels for various age groupings")
```

# Conduct Statistical Significance Testing

Statistical significance testing is critical to understanding whether
estimates are meaningfully different. Estimates that may appear
substantially different in isolation are frequently, especially at
smaller geographies, not statistically significantly different because
their errors are so significant.

We'll illustrate this numerically and demonstrate the impacts of
accounting for error when visualizing data, leveraging
`tidycensus::significance()` to conduct our actual tests. Here, we'll
compare each tract-level value within a single county to that of the
corresponding county.

```{r, fig.width = 7}
plot_data = acs_df_tract %>%
  filter(str_detect(NAME, "Atlantic")) %>%
  select(GEOID, matches("age_over_64_percent")) %>%
  mutate(county_geoid = str_sub(GEOID, 1, 5)) %>%
  left_join(
    acs_df_county %>%
      filter(str_detect(NAME, "Atlantic")) %>%
      select(GEOID, matches("age_over_64_percent")) %>%
      rename_with(
        .cols = matches("age_over_64_percent"),
        .fn = ~ str_c(.x, "_county")),
    by = c("county_geoid" = "GEOID")) %>%
  mutate(
    naive_difference = case_when(
      age_over_64_percent > age_over_64_percent_county ~ "Larger",
      age_over_64_percent < age_over_64_percent_county ~ "Smaller",
      TRUE ~ "Equal"),
    significance = significance(
      est1 = age_over_64_percent,
      est2 = age_over_64_percent_county,
      moe1 = age_over_64_percent_M,
      moe2 = age_over_64_percent_M_county,
      clevel = 0.9),
    statistical_difference = case_when(
      significance == FALSE ~ "Not significant",
      significance == TRUE & age_over_64_percent > age_over_64_percent_county ~
        "Larger",
      significance == TRUE & age_over_64_percent < age_over_64_percent_county ~
        "Smaller"))

plot_data %>%
  select(GEOID, naive_difference, statistical_difference) %>%
  pivot_longer(cols = -c(GEOID, geometry)) %>%
  mutate(
    name = if_else(
      name == "naive_difference",
      "Difference in point estimates",
      "Statistically-significant difference")) %>%
  ggplot() +
    geom_sf(aes(fill = value)) +
    scale_fill_manual(
      values = c(
        "Larger" = palette_urbn_main[1] %>% as.character(),
        "Smaller" = palette_urbn_main[2] %>% as.character(),
        "Not significant" = palette_urbn_main[4] %>% as.character())) +
    theme_urbn_map() +
    facet_wrap(~ name) +
    theme(legend.position = "bottom", legend.direction = "horizontal") +
    labs(
      fill = "",
      title = "Many Tract-Level Estimates Are Not Statistically Significantly Different from the County-Level Estimate" %>% str_wrap(90),
      subtitle = "Tract-level share of the population 65+ vs. county-level estimate, Atlantic County, NJ")

```

```{r}
significance_test_data = plot_data %>%
  filter(GEOID %in% c("34001011901", "34001001900"))

## TRUE
significant_difference = significance(
  est1 = significance_test_data %>%
    filter(GEOID == "34001001900") %>%
    pull(age_over_64_percent),
  est2 = significance_test_data %>%
    filter(GEOID == "34001011901") %>%
    pull(age_over_64_percent),
  moe1 = significance_test_data %>%
    filter(GEOID == "34001001900") %>%
    pull(age_over_64_percent_M),
  moe2 = significance_test_data %>%
    filter(GEOID == "34001011901") %>%
    pull(age_over_64_percent_M))

plot_data %>%
  filter(GEOID %in% c("34001011901", "34001001900")) %>%
  ggplot(aes(y = GEOID, x = age_over_64_percent, color = GEOID)) +
    geom_point() +
    geom_errorbar(
      aes(
        xmin = age_over_64_percent - age_over_64_percent_M,
        xmax = age_over_64_percent + age_over_64_percent_M),
      width = 0.2) +
  theme_urbn_print() +
  labs(
    x = "Share of population over 64",
    y = "Tract",
    title = "One Tract Estimate Doubles That of the Other, But There is No Statistically Significant Difference" %>% str_wrap(100),
    subtitle = "Share of population over the age of 64, point estimates and margins of error")
```

# Report and Visualize Error

The ability to quantify error both allows us to conduct statistical
significance testing and to directly report error, either in text and/or
via visualizations. As illustrated above, MOEs can be reported directly
alongside estimates in the text as parentheticals: estimate (±MOE). A
similar approach can be employed for statistical significance testing,
with statistical significance being notated as a "\*", for example, or
via a longer "p \< .10" statement.

For most such cases, it will be helpful to provide some guidance to your
audience about what the MOE/statistical significance means and how it
should shape their interpretation of point estimates, such as via a
footnote/endnote, through a call-out box, or in a section on your
methods.

In tables, you can either present MOEs/statistical significance in the
same cell as the point estimates or via a companion column. Depending on
your goals, CVs may be used to suppress altogether some estimates within
a table, though this approach also has limitations (e.g., it may lead to
obscuring important but small-population groups or geographies, not just
in a single table but across many such tables). If you're struggling
with presenting results because many estimates are imprecise, this may
be an indication that you need to re-evaluate your choice of geography
and/or explore opportunities for collapsing variables to obtain more
precise estimates.

There are myriad approaches to visualizing error. Generally, the use of
error bars (look no further than the visual immediately above) should be
avoided, as research consistently shows that most audiences do not
interpret such bars accurately. The excellent `library(ggdist)` provides
an array of methods for different styles and methods of visualizing
error, while visualizing statistical significance relative to a
benchmark (see the map above) is another common approach.

# Limitations of Calculated Errors

MOEs for raw ACS estimates–i.e., those that are directly reported by the
Census Bureau–should be accurate measurements of survey error. This is
because these errors are calculated after accounting for estimate
covariance using individual-level responses (see footnote 1). However,
the calculations of error for derived estimates–that is, estimates that
`library(urbnindicators)` calculates behind the scenes using two or more
raw ACS estimates are imperfect for at least two reasons:

1.  Errors for derived estimates do not account for covariance;

2.  Errors for derived estimates are calculated using (Census
    Bureau-recommended) formulae that inherently inflate error when
    multiple variables are combined. For example, `age_over_64_percent`
    (see the first figure above) is calculated by summing numerous raw
    estimates to produce the numerator and then dividing this numerator
    by the table universe. As shown, this process produces a much more
    precise estimate than is available for any of the component variables
    used to calculate the numerator–but the error for this newly-derived
    estimate is likely larger than the directly-calculated (and more
    accurate) error when using individual-level data.

***So what to do?***

The Census Bureau produces a set of auxiliary [variance replicate
estimate
tables](https://www.census.gov/programs-surveys/acs/data/variance-tables.html)
(VRE) that allow users to calculate precise errors akin to those
calculated using individual-level responses. Users focused on the most
precise methods should consider whether these VRE tables are appropriate
for their needs. Unfortunately, these tables are only produced for a
small subset of all ACS tables, and the VRE tables are large and
non-trivial to work with.

A similar option that may be appropriate for some use-cases is to work
with the the individual-level data, which is often referred to as the
"Public Use Microdata". These data comprise respondent-level ACS data as
well as large set of survey replicate weights that can be used to
calculate more accurate margins of error and related error statistics.
However, these data are much more cumbersome to work with than the
tabulated estimates tables returned by `urbnindicators`, and in order to
protect respondent anonymity, the microdata can only produce estimates
at geographies of \~100,000 people or more (mid-sized cities and
counties, as well as special statistical geographies called Public Use
Microdata Areas, or PUMAs).

If the tabulated estimates remain the best solution for your use case,
the codebook that is returned alongside the data from
`compile_acs_data()` can inform users' understandings of the number of
raw estimates used to calculate a given variable, which in turn can
shape their thinking around the precision or possible imprecision of
associated errors.
