---
title: "urbnindicators"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{urbnindicators}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  markdown: 
    wrap: 72
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  warning = FALSE,
  message = FALSE,
  collapse = TRUE,
  dpi = 1000,
  comment = "#>",
  out.width = "100%")
```

```{r setup, echo = FALSE}
library(urbnindicators)
library(tidycensus)
library(dplyr)
library(stringr)
library(ggplot2)
library(scales)
library(urbnthemes) # install from https://github.com/UrbanInstitute/urbnthemes
```

This vignette is organized into three parts:

1.  We illustrate a typical workflow for obtaining American Community
    Survey (ACS) data using `library(tidycensus)`.

2.  We show how `library(urbnindicators)` can produce similar outputs, but with
    much less code and effort.

3.  We highlight how `library(urbnindicators)` provides metadata and improves the
    quality and reliability of the analysis process.

## A Typical Workflow

[**tidycensus**](https://walker-data.com/tidycensus/index.html) provides
a suite of functions for working with ACS data in R. While it's versatile 
and comprehensive, it can require a significant amount of knowledge and
time to support a robust analysis, leading many users to fall into
common pitfalls without realizing they've made an error(s).

### Identify variables to query

We load the built-in codebook and search for our construct of interest
(disability). This leaves us 500 variables to choose from.

```{r}
acs_codebook = load_variables(dataset = "acs5", year = 2022)

acs_codebook %>%
  filter(str_detect(concept, "Disability")) %>% 
  select(name, label, concept) %>%
  ## only printing three, for brevity
  head(3) %>%
  reactable::reactable()
```

Let's imagine we're interested in calculating the share of individuals
with a disability; this requires only two variables (conceptually): the
number of people with a disability and the number of all people.

So which variable(s) do we select? There's not a clear answer. All
variables relating to disability reflect disability and at least one
other characteristic (e.g., "sex by age by disability status"). If we
want to calculate the percent of all individuals with a disability, we
want to do so using the most robust available variables (i.e., those
that reflect all individuals who reported their disability status),
whereas some variables that reflect disability may have smaller counts
because the other characteristics combined with disability status (e.g.,
health insurance coverage status) may be available only for a subset of
the individuals for whom disability status is available.

Let's imagine we select the table of variables prefixed "B18101", 
for "Sex by Age by Disability". We think that most respondents who respond 
about their disability status will also have responded about their sex and 
age. We then pass this to `library(tidycensus)` as:

```{r}
df_disability = get_acs(
  geography = "county",
  state = "NJ", 
  year = 2022,
  output = "wide",
  survey = "acs5",
  table = "B18101")
```

This returns us 21 observations–one for each county in NJ--along with an
intimidating 80 columns with unintelligble names along the lines of
`B18101_039E`.

### Calculating our measure of interest

Now we would need to figure out how to aggregate the needed variables
for both the denominator and numerator in order to calculate a valid "%
Disabled" measure, a task that is feasible but time-intensive and
error-prone.

For an analysis that leverages more than a single measure, and
especially when measures are required from distinct tables, this
workflow is burdensome and creates significant surface area for
undetected errors.

At the same time, many analysts will be overwhelmed by and unsure how to
combine the margins of error that are returned by `tidycensus::get_acs()`
to calculate pooled errors for the new percent-disabled variable, opting
simply to drop this critical information from their analysis. (See 
(quantified-survey-error)[quantified-survey-error.html] to learn more about 
how `library(urbnindicators)` helps simplify this task.)

## Using urbnindicators

`library(urbnindicators)` abstracts the workflow above behind the scenes. Instead
of a call to `tidycensus::get_acs()`, a call to
`urbnindicators::compile_acs_data()` returns a dataset of both raw ACS
measures and derived estimates (such as the share of all individuals who
are disabled). And that dataset can include a range of measures–-spanning
things such as health insurance, employment, housing costs, and race and
ethnicity–-not just one variable or table from the ACS.

### Acquire data

It's as simple as the call below. Note that you can provide a vector of
years and/or states if you want data over different time periods or
geographies.

Note that selecting more tables or more geographic units--either by selecting 
a `geography` option comprising more units, by selecting more states, or selecting 
more years--can significantly increase the query time. A tract-level query of the 
entire US for all supported variables can take 30+ minutes.

Use `list_tables()` to see some of the most commonly-used tables:

```{r}
list_tables() |> head(10)
```

Or use `get_acs_codebook()` to see every table supported by the Census Bureau API:

```{r}
get_acs_codebook() |>
  ## just showing a sample of the 28,000+ variables available
  slice_sample(n = 10)
```

Here we request just two tables--`disability` and `transportation_to_work`.

```{r, message = FALSE, warning = FALSE}
df_urbnindicators = compile_acs_data(
  years = 2024,
  tables = c("disability", "transportation_to_work"),
  geography = "county",
  states = "NJ",
  spatial = TRUE)
```

Alternately, you can pass the name of a variable or table from `get_acs_codebook()`
to `compile_acs_data()`. The equivalent would be:

```{r}
df_urbnindicators = compile_acs_data(
  years = 2024,
  tables = c("sex_by_age_by_disability_status_universe", "B08301"),
  geography = "county",
  states = "NJ",
  spatial = TRUE)
```

### Analyze or visualize data

And now we're ready to analyze or plot our data. Simplistically:

```{r}
df_urbnindicators %>%
  ggplot() +
    geom_sf(aes(fill = disability_percent)) +
    theme_urbn_map() +
    scale_fill_continuous(labels = scales::percent, transform = "reverse") +
    labs(
      title = "Disability Rates Appear Higher in Southern NJ",
      subtitle = "Disability rates by county, NJ, 2018-2022 ACS",
      fill = "Population with an ACS-defined disability (%)" %>% str_wrap(20))
```

### Document data

There's a lot happening behind the scenes, so it's important to
understand what each variable represents and how it was calculated.
`library(urbnindicators)` includes a codebook as an attribute of the dataframe
returned from `urbnindicators::compile_acs_data()`. View and navigate through the 
[full codebook here](articles/codebook.html).

The codebook specifies the variable type and provides a definition of
how the variable was calculated. Most (though not all) variables that
are directly available from the ACS are count variables. Many of the
variables that are calculated by `library(urbnindicators)` are percent
variables, where we divide two count variables. 

Others, however, are quite complex. For example, `disability_percent` is 
the sum of all of the sex-by-age groupings for people with disabilities (numerator)
divided by the table universe.

```{r}
df_urbnindicators %>%
  attr("codebook") %>%
  filter(str_detect(calculated_variable, "^disability_percent$")) %>%
  pull(definition)
```
