---
title: "Explore ACS Data Interactively"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore ACS Data Interactively}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(urbnindicators)
```

`view_acs_data()` opens an interactive dashboard for exploring an ACS dataset:
a choropleth map, a searchable table of measures, on-the-fly benchmarking
against parent geographies, and tools for interpolating to custom areas.

```{r demo-video, echo = FALSE, results = "asis"}
# The demo video is shown only on the pkgdown website, not in the installed
# package vignette. It lives in pkgdown/assets/ (which is in .Rbuildignore, so
# it never enters the package build); pkgdown copies that directory to the site
# root and serves the file standalone, so it is referenced -- not embedded --
# keeping the article lightweight. pkgdown sets IN_PKGDOWN = "true" while
# building the site.
if (identical(Sys.getenv("IN_PKGDOWN"), "true")) {
  cat(
    "The short demo below walks through the interface before we build one ",
    "ourselves.\n\n",
    '<video controls playsinline preload="metadata" ',
    'style="width: 100%; max-width: 960px; height: auto; display: block; ',
    'margin: 1rem auto; border: 1px solid #e0e0e0; border-radius: 4px;">\n',
    '  <source src="../view_acs_data-demo.mp4" type="video/mp4">\n',
    '  Your browser does not support embedded video. ',
    '<a href="../view_acs_data-demo.mp4">Download the demo</a>.\n',
    '</video>',
    sep = "")
}
```

```{r}
df1 = compile_acs_data(
  tables = "cost_burden",
  states = "CA",
  counties = "06037",
  years = "2024",
  geography = "tract", 
  spatial = TRUE)

places1 = tigris::places(state = "NJ", cb = TRUE, year = 2024) %>%
  sf::st_filter(df1)
```


The `geography_extent` argument is required: it declares which higher geographies
are valid *statistical benchmarks* for these data. Because `df1` was pulled for a
single county (`counties = "06037"`, Los Angeles County), only a **county**
benchmark is appropriate -- the tracts here don't cover all of California, so a
state benchmark would compare each tract to a partial, misleading "state" value.
Had we pulled the full state (`counties = NULL`), we could declare
`geography_extent = c("county", "state")`. Pass `geography_extent = "none"` to
turn benchmarking off entirely.

```{r}
view_acs_data(df1, geography = "tract", geography_extent = "county")
```

## Custom target geographies

`view_acs_data()` can also interpolate the source data onto a user-supplied
polygon set (e.g., neighborhoods, council districts, school catchments) via
area-weighted overlap. Pass an `sf` object to `target_geographies`; it must
contain a `GEOID` column that uniquely identifies each target polygon. A
"Geography" toggle then appears in the sidebar so you can switch between the
source view and the interpolated target view.

Benchmark values continue to be computed from the source data. In Target view,
each target polygon is mapped to a parent (county / state / national) by
majority-area overlap of its source-geography components.

Any `sf` polygon layer with a unique `GEOID` works. Here we use Census
"places" (incorporated cities and towns) as the target geographies; an
optional `NAME` column, when present, is shown in the map popups:

```{r, eval = FALSE}
view_acs_data(
  df1,
  geography          = "tract",
  geography_extent   = "county",
  target_geographies = places1)
```

Under the hood, this builds a fractional crosswalk via
`sf::st_intersection()` and feeds it to [`interpolate_acs()`].

## Draw your own geographies

You don't need a target layer in hand. Every map includes a drawing toolbar at
its top-left: draw one or more polygons, then click **Interpolate to drawn
area** in the sidebar. The source data are interpolated onto your polygons with
the same area-weighted crosswalk, the map switches to **Target** view, and
benchmarking (when available) compares each drawn area to the parent geography
it mostly overlaps. **Clear target** discards the polygons and returns you to
the source view (or to a `target_geographies` layer, if you supplied one).