---
title: "Explore ACS Data Interactively"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Explore ACS Data Interactively}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(urbnindicators)
```


```{r}
df1 = compile_acs_data(
  tables = "cost_burden",
  states = "NJ",
  counties = "34013",
  year = "2024",
  geography = "tract", 
  spatial = TRUE)
```


```{r}
view_acs_data(df1, geography = "tract")
```

## Custom target geographies

`view_acs_data()` can also interpolate the source data onto a user-supplied
polygon set (e.g., neighborhoods, council districts, school catchments) via
area-weighted overlap. Pass an `sf` object to `target_geographies`; it must
contain a `GEOID` column that uniquely identifies each target polygon. A
"Geography" toggle then appears in the sidebar so you can switch between the
source view and the interpolated target view.

Benchmark values continue to be computed from the source data. In Target view,
each target polygon is mapped to a parent (county / state / national) by
majority-area overlap of its source-geography components.

```{r, eval = FALSE}
neighborhoods = sf::st_read("path/to/neighborhoods.geojson")

view_acs_data(
  df1,
  geography          = "tract",
  target_geographies = neighborhoods)
```

Under the hood, this builds a fractional crosswalk via
`sf::st_intersection()` and feeds it to [`interpolate_acs()`].